---
title: "CU traits_Saliva"
author: "Elena"
date: "2024-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
library(tidyverse)
library(foreign)
library(lme4)
library(car)
library(broom)
library(broom.mixed)
library(knitr)
library(kableExtra)
library(xfun)
data <- read.spss("Data/CEDS_Archived.sav", to.data.frame=TRUE)
```

Data wrangling
```{r}
#Select variables
data_selected<- data %>% select(ID, Saliva1Time, Saliva2Time, Saliva3Time, Saliva4Time, Saliva5Time, Saliva6Time, Saliva1done, Saliva2done, Saliva3done, Saliva4done, Saliva5done, Saliva6done, ICU03, ICU05, ICU13, ICU15, ICU16, ICU17, ICU23, ICU24, ICU02, ICU04, ICU07, ICU08, ICU09, ICU10, ICU11, ICU12, ICU18, ICU20, ICU21, ICU01, ICU06, ICU14, ICU19, ICU22, cdspa) %>% mutate(ID = as.factor(ID))

#Conduct disorder - classify people with or without conduct disorder based on the standardized scores provided in the cdspa variable. People that have 0 are classified as "no" (CD), people that have 1 or more are classified as "yes" (CD)."yes" - 1, "no" - 0

data_CD<- data_selected %>% mutate(CD = case_when(cdspa == 0 ~ "0",
                                                       TRUE ~ "1"))%>% mutate(CD = as.factor(CD))

#Re-code CU variables

data_gathered<- data_CD %>% gather(Item, Score, ICU03:ICU22) %>% mutate(Score = dplyr::recode(Score, 'not at all true' = '0', 'somewhat true' = '1', 'very true' = '2', 'definitely true' = '3'))%>% mutate(Score = as.numeric(Score))

#Some items are reverse-coded

data_reversed <- data_gathered %>% mutate(Score=if_else(Item =='ICU03'| Item =='ICU05' | Item =='ICU13' | Item =='ICU15'|Item =='ICU16'| Item =='ICU17' | Item =='ICU23' | Item =='ICU24' | Item == 'ICU08' | Item == 'ICU01' | Item == "ICU14" | Item == "ICU19", 3-Score,Score))
```

Check missingness in CU trait questions and create sum scores
```{r}
data_CU_filtered<- data_reversed %>% filter(!is.na(Score))
data_summed<- data_CU_filtered %>% group_by(ID) %>% mutate(sum_score = sum(Score))%>% filter(!is.na(sum_score)) #remove participants that do not have a sum score
```

Gather and check missingness in saliva data
```{r}
#Gather saliva / time

data_gather_saliva<- data_summed %>% gather(Time, Saliva, Saliva1Time:Saliva6Time) %>% gather(Completion, Status, Saliva1done: Saliva6done)

#recode completion/ status data to check missingness
#give 1 for completed, 0 for not completed

data_gather_saliva<- data_gather_saliva %>% mutate(Status = case_when(Status == "yes" ~ 1,
                                                             TRUE ~ 0))
data_gather_saliva %>% group_by(Completion)%>% summarise(sum = sum(Status))

#Attempting to remove all those that do not have completed data

data_gather_saliva<- data_gather_saliva %>% filter(Status == 1) #include only participants that have completed all 7 saliva samples.

#Check the minimum and maximum CU trait score
min(data_gather_saliva$sum_score) #0 
max(data_gather_saliva$sum_score) #56

#We use the mean +/- SD to clasify participants into high, medium and low CU trait groups
```

Summary Stats overall no groupings
```{r}
#Summary Statistics overall no groupings

summary_CU <-  data_gather_saliva%>% ungroup()%>%
  summarise(n=n(),
            mean=mean(sum_score, na.rm = T),
            sd=sd(sum_score, na.rm = T),
            se=sd/sqrt(n),
            lower=mean-se*1.96,
            upper=mean+se*1.96)

summary_CU

#Mean score is 17.12512	, sd is 8.643162	

#X - axis is minutes since session initiation, y axis is cortisol, lines groups by high (+1 SD), Average (mean), and low (-1SD)

#Group by high low baseline

data_grouped<- data_gather_saliva %>% mutate(Grouping = case_when(sum_score > 25.76828 ~ "High",
                                                             sum_score < 8.481958 ~ "Low",
                                                             TRUE ~ "Medium"))%>% mutate(Time = as.factor(Time))


data_grouped$Grouping <- factor(data_grouped$Grouping, levels=c('Low', 'Medium', 'High'))
```
Hypothesis 1: There will be a group difference, i.e., conduct disorder or not, in terms of how the cortisol levels change over time. 

Making plots for conduct disorder (CD) status

```{r}
summary_Time_CD <-  data_grouped %>% group_by(Time,CD) %>%
  summarise(n=n(),
            mean=mean(Saliva, na.rm = T),
            sd=sd(Saliva, na.rm = T),
            se=sd/sqrt(n),
            lower=mean-se*1.96,
            upper=mean+se*1.96)%>%tibble()
summary_Time_CD

CD_status_plot<- ggplot(summary_Time_CD, aes(x = Time, y = mean, group = CD)) +
  geom_line(aes(lty = as.character(CD), color = CD),
            position = position_dodge(width = 0.1)) +
  geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1,
                position = position_dodge(width = 0.1)) +
  geom_point(aes(shape = as.character(CD)),
             position = position_dodge(width = 0.1)) +
  guides(lty = "none", shape = "none")+
  theme_classic()+
  labs(x="Sample Time",y="Cortisol Level")+
scale_x_discrete(
    "Sample Time",
    labels = c(
      "Saliva1Time" = "20mins_prior",
      "Saliva2Time" = "10mins_prior",
      "Saliva3Time" = "End_Task",
      "Saliva4Time" = "20mins_after",
      "Saliva5Time" = "45mins_after",
      "Saliva6Time" = "60_mins_after"))+
  scale_color_discrete(
  labels = c("No", "Yes"),
  name = "Conduct Disorder Status")

ggsave("Plots/CD_status_plot.png", plot = CD_status_plot, width = 8, height = 6, dpi = 300)
```
Statistical Analysis (H1)

```{r}
data_grouped$CD <- factor(data_grouped$CD, levels=c('0', '1'))

contrasts(data_grouped$CD) <- c(1/2, -1/2)

data_grouped$Time <- factor(data_grouped$Time)

contrasts(data_grouped$Time) <- contr.poly(6) #polynomial contrast codes 

mod_H1 <- afex::mixed(Saliva ~ Time*CD + (1|ID), data=data_grouped, check_contrasts = FALSE)
summary(mod_H1)
anova(mod_H1, type=3)

#Test assumptions for linear mixed effect models

tdat <- data.frame(predicted=predict(mod_H1$full_model), residual = residuals(mod_H1$full_model))

ggplot(tdat,aes(x=predicted,y=residual)) + geom_point() + geom_hline(yintercept=0, lty=3)

ggplot(tdat,aes(x=residual)) + geom_histogram(bins=40, color="black")

ggplot(tdat,aes(sample=residual)) + stat_qq() + stat_qq_line()

raneff <- lme4::ranef(mod_H1$full_model)

hist(raneff$ID$`(Intercept)`, breaks=15) #random effect of participant normal
```

H2. Individuals in the high CU traits in the CD group will have a less steep change over time in cortisol levels.

```{r}
#YES CD group only - much greater variability

data_CD_yes<- data_grouped %>% filter(CD == "1")

data_CD_yes$Time <- as.numeric(data_CD_yes$Time)
data_CD_yes$Saliva <- as.numeric(data_CD_yes$Saliva)

# Ensure 'ID' and 'CD' are factors or characters
data_CD_yes$ID <- as.factor(data_CD_yes$ID)
data_CD_yes$CD <- as.factor(data_CD_yes$CD)

# Remove rows with NA values in 'Time' or 'Saliva'
data_CD_yes <- na.omit(data_CD_yes)

#library(ggplot2)

ggplot(data_CD_yes, aes(x = Time, y = Saliva, colour = ID)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "none") +
  expand_limits(x = range(data_CD_yes$Time))+
  labs(x="Sample Time",y="Cortisol Level")+
scale_x_continuous(breaks=c(1,2,3,4,5,6))
```


```{r}
#NO CD group only

data_CD_no<- data_grouped %>% filter(CD == "0")

data_CD_no$Time <- as.numeric(data_CD_no$Time)
data_CD_no$Saliva <- as.numeric(data_CD_no$Saliva)

# Ensure 'ID' and 'CD' are factors or characters
data_CD_no$ID <- as.factor(data_CD_no$ID)
data_CD_no$CD <- as.factor(data_CD_no$CD)

# Remove rows with NA values in 'Time' or 'Saliva'
data_CD_no <- na.omit(data_CD_no)

library(ggplot2)

ggplot(data_CD_no, aes(x = Time, y = Saliva, colour = ID)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "none") +
  expand_limits(x = range(data_CD_yes$Time))+
  labs(x="Sample Time",y="Cortisol Level")+
scale_x_continuous(breaks=c(1,2,3,4,5,6))
```

Plots including CU traits

```{r}
summary_Time_Grouping_CD <-  data_grouped %>% group_by(Time, Grouping, CD) %>%
  summarise(n=n(),
            mean=mean(Saliva, na.rm = T),
            sd=sd(Saliva, na.rm = T),
            se=sd/sqrt(n),
            lower=mean-se*1.96,
            upper=mean+se*1.96)
summary_Time_Grouping_CD%>%filter(CD == 1)

custom_labels <- c("0" = "No CD", "1" = "Yes CD")

ggplot(summary_Time_Grouping_CD, aes(x = as.numeric(Time), y = mean, group = Grouping)) +
  geom_line(aes(lty = as.character(Grouping), color = Grouping),
            position = position_dodge(width = 0.1)) +
  geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1,
                position = position_dodge(width = 0.1)) +
  geom_point(aes(shape = as.character(Grouping)),
             position = position_dodge(width = 0.1)) +
  guides(lty = "none", shape = "none")+
  theme_classic()+
  labs(x="Sample Time",y="Saliva")+
  facet_wrap(~ CD, labeller = labeller(CD = custom_labels))+
scale_x_continuous(breaks=c(1,2,3,4,5,6,7))
```

Statistical Analysis (H2)
```{r}
data_CD_yes$Time <- factor(data_CD_yes$Time)

contrasts(data_CD_yes$Time) <- contr.poly(6) #polynomial contrast codes 

contrasts(data_CD_yes$Grouping) <- cbind(c1 =  c(2/3, -1/3, -1/3),
                                 c2 = c(1/2, 0, -1/2))
contrasts(data_CD_yes$Grouping)

mod_H2 <- afex::mixed(Saliva ~ Time + (1|ID), data=data_CD_yes, check_contrasts = FALSE)
summary(mod_H2)
anova(mod_H2, type=3)

mod_H2_group <- afex::mixed(Saliva ~ Time*Grouping + (1|ID), data=data_CD_yes, check_contrasts = FALSE)
summary(mod_H2_group)
anova(mod_H2_group, type=3)

anova(mod_H2, mod_H2_group) #significant

#Test assumptions for linear mixed effect models

tdat <- data.frame(predicted=predict(mod_H2$full_model), residual = residuals(mod_H2$full_model))

ggplot(tdat,aes(x=predicted,y=residual)) + geom_point() + geom_hline(yintercept=0, lty=3)

ggplot(tdat,aes(x=residual)) + geom_histogram(bins=40, color="black")

ggplot(tdat,aes(sample=residual)) + stat_qq() + stat_qq_line()

raneff <- lme4::ranef(mod_H2$full_model)

hist(raneff$ID$`(Intercept)`, breaks=15) #random effect of participant normal
```


